---
title: Programming in R
subtitle: R Foundations course
author: Ella Kaye | Department of Statistics | University of Warwick
title-slide-attributes:
  data-background-color: "#552D62"
date: 2022-10-31
date-format: long
format: 
  revealjs:
    theme: [default, ../../slides.scss]
    slide-number: true
    execute:
      echo: true
      code-overflow: wrap
---


## Overview

- More on data structures

- Control flow and iteration functions

- Efficient R programming

- Writing functions (basics)

## Intro

Understanding the basics of R programming helps to improve
analysis/reporting scripts and extend what we can do with R.

Good coding practice follows the DRY principle: **D**on't **R**epeat **Y**ourself. 
Rather than modifying copy-pasted code chunks, we might

- write a custom function
- use loops or iteration functions to perform multiple similar tasks
 
Custom functions can be used to provide convenient wrappers to complex 
code chunks as well as implement novel functionality.


# More on Data Structures {background-color="#552D62"}

## Data Structures Revisited

For basic data analysis, our data is usually imported and we use 
high-level functions (e.g. from **dplyr**) to handle it.

For programming, we need to work with lower-level data structures and be
able to

- create basic objects
- extract components
- coerce one data type to another

Working with base R functions when programming also helps avoid dependencies, which is useful when writing packages.


## Vectors

`numeric()`, `character()` and `logical()` can be used to initialize 
vectors of the corresponding type for a given length

```{r}
x <- numeric(3)
x
```

Elements can be assigned by indexing the positions to be filled, e.g.

```{r}
x[1] <- 4 # assign 4 to 1st element
x[-c(2, 3)] <- 4 # assign 4 to everying *except* 2nd and 3rd element
```

This is particularly useful when programming an iterative procedure.

`as.logical()`, `as.numeric()` and `as.character()` coerce to the 
corresponding type, producing `NA`s if coercion fails.

## Logical Vectors

Logical vectors are commonly used when indexing. The vector might be 
produced by a logical operator:
```{r}
x <- c(1, 1, 2, 2, 2)
x > 1
x[x > 1]
```
`duplicated()` is also useful here:
```{r}
duplicated(x)
!duplicated(x)
```

## Numeric Vectors

The are several convenience function for creating
numeric vectors, notably `seq()` and `rep()`.

As they are so useful there are fast shortcuts for
particular cases
```{r}
seq_len(4)
fruits <- c("apple", "pear", "banana")
seq_along(fruits) # a sequence from 1 to the length of x
rep.int(1:2, times = c(2, 3))
```

## Character Vectors

Character vectors may be used for creating names
```{r}
x <- 3:5
names(x) <- paste0(LETTERS[1:3], 1229:1231)
x
names(x)
```
Names can be used as an alternative to numeric or logical vectors when indexing
```{r}
x["B1230"]
```

## Matrices

A matrix is in fact also a vector, with an attribute giving the dimensions of the matrix

::: {.panel-tabset}
## default
```{r}
M <- matrix(1:6, 2, 3) # data, nrow, ncol
M
str(M)
attributes(M)
```

## byrow
The `byrow` argument is also useful:

```{r}
N <- matrix(1:6, 2, 3, byrow = TRUE) # data, nrow, ncol
N
```
:::

## Matrix functions 

Useful functions for matrices include `dim()`, `ncol()`, `nrow()`, 
`colnames()` and `rownames()`. `rbind()` and `cbind()` can be used to 
row-bind or column-bind vectors.

Matrices enable computation via matrix algebra as well as row/column-wise operations.

## Lists

Lists collect together items which may be different types or lengths. Like a 
vector, elements may be named.
```{r}
results <- list(matrix = M, vector = x)
results
```

Lists are often used to return the results of a function. 

## Indexing lists

Elements can be indexed by `[` to return a list or `[[` to return a 
single element, either by index or name:

```{r}
results["vector"] # returns a list
results[[2]] # returns a vector
```


`$` can be used to extract elements by name:

```{r}
results$vector # equivalent to results[["vector"]]
```

## Data Frames

Data frames are lists of variables of the same length and hence can often 
be treated as a matrix 
```{r}
x <- 1:3
dat <- data.frame(x = x, id = letters[1:3])
dat
dat[1]
dat[[1]]
dat[1, 2]
```



## Your turn!

The `lm` function calls the "workhorse" function `lm.fit` to actually 
fit the model. Unlike `lm`, which works from a formula, `lm.fit` works 
from the model matrix and the response vector.

1. Define a response `y` containing 10 numeric values. Define an 
explanatory variable `z` of the numbers 1 to 10.

1. Use the function `cbind()` to create a matrix `x` with 1s in the first
column and `z` in the second column.

1. Fit a model using `fit1 <- lm.fit(x, y)`. Use `str` to explore the structure of
the results. Use `$` to extract the coefficients.

1. Create a second fit using `lm(y ~ z)`. Use `names` to compare the 
results. Check the coefficients of the second fit are the same. 

## Your turn! (a solution)

```{r}
#| eval: false

# A possible solution

set.seed(1)
y <- sample(10)
z <- 1:10

x <- cbind(1, z) # we'll see this trick in a few slides time!

fit1 <- lm.fit(x,y)
str(fit1)
fit1$coefficients

fit2 <- lm(y ~ z)
names(fit1)
names(fit2)
fit2$coefficients
```


# Control Flow {background-color="#552D62"}

## Control Structures

Control structures are the commands that make decisions or execute loops.

Conditional execution: `if`/`else`, `switch`

Loops: `for`, `while`, `repeat`


## `if`/`else`

An `if` statement can stand alone or be combined with an `else` statement

```{r}
x <- 1:3
if (all(x > 0)) {
    res <- mean(x)
} else {
    res <- mean(abs(x))
}
res
```

The condition must evaluate to logical vector of length one. The 
functions `all()`, `any()`, `is.na()`, `is.null()` and other `is.`
functions are useful here.

## Conditioning on equality

Using `==` may not be appropriate as it compares each element; `identical()` will test the whole object

```{r}
x <- y <- 1:2
x == y
identical(x, y)
```

`all.equal()` will allow for some numerical “fuzz”

```{r}
z <- sqrt(2)
identical(z * z, 2)
all.equal(z * z, 2)
```

## `switch`

The `switch()` function provides a more readable alternative to nested 
`if` statements
```{r}
#| eval: false
if (summary == "IQR") { 
    y <- IQR(x)
} else {
    if (summary == "range"){
        y <- range(x)
    } else y <- mean(x)
}
```


```{r}
x <- 1:5
switch("range", # can enter an arg name or position
       IQR = IQR(x),
       range = range(x),
       mean(x))
```

The final unnamed argument is the default.

[Further examples](https://www.datamentor.io/r-programming/switch-function/){target="_blank}

## `for`

A for loop repeats a chunk of code, iterating along the values of a 
vector or list

```{r}
x <- c("apple", "pear")
for (nm in x) print(nm)
```

Unassigned objects are not automatically printed; hence call to `print()`. 
This also applies to `ggplot` objects, which only display when printed.

```{r}
for (i in seq_along(x)) {
    message("Element ", i, ": ", x[i])
}
```

`seq_along()` is used here rather than `1:length(x)` as `length(x)` may 
be zero. `message` is used to print messages to the console.

## `while` and `repeat`

The `while` loop repeats while a condition is `TRUE`
```{r}
#| eval: false
n_iter <- 1
while (n_iter < 3) {
    x <- x * 2
    n_iter <- n_iter + 1
}
```

The `repeat` loop repeats until exited by `break`
```{r}
#| eval: false
repeat {
    x <- x + 1
    if (max(x) > 10) break
}
```

`break` can be used in `for` or `while` loops too. 

`next` can be used to skip to the next iteration.


## Iteration functions

Iteration functions provide a general alternative to for loops. They are 
not necessarily faster, but can be more compact.

`apply()` applies a function over rows/columns of a matrix.

`lapply()`, `sapply()` and `vapply()` iterate over a list or vector. `vapply()` 
is recommended for programming as it specifies the type of return value
```{r}
vapply(list(a = 1:3, b = 1:6), FUN = mean, FUN.VALUE = numeric(1))
```
`mapply()` iterates over two or more lists/vectors in parallel.

## Iteration function resources

- [Efficient R](https://csgillespie.github.io/efficientR/programming.html#the-apply-family){target="_blank"} by Colin Gillespie and Robin Lovelace

- The built-in help pages. You can directly access the examples using the `example()` function, e.g. to run the `apply()` examples, use `example("apply")`.

- This StackOverflow [answer](https://stackoverflow.com/questions/3505701/grouping-functions-tapply-by-aggregate-and-the-apply-family){target="_blank"}, describing when, where and how to use each of the functions. 

- This blog [post](https://nsaunders.wordpress.com/2010/08/20/a-brief-introduction-to-apply-in-r/){target="_blank"} by Neil Saunders


## **purrr**

The [**purrr**](https://purrr.tidyverse.org){target="_blank"} package (part of the tidyverse) provides alternatives to the apply family that have a simpler, more consistent interface with fixed type of return value.

```{r}
# Split a data frame into pieces, 
# fit a model to each piece, summarise and extract R^2
library(purrr)

mtcars %>%
  split(.$cyl) %>% # base R
  map(~ lm(mpg ~ wt, data = .x)) %>% # returns a list
  map(summary) %>%
  map_dbl("r.squared") # returns a vector
```

## Advantages of **purrr**

- The first argument is always the data, so **purrr** works naturally with the pipe.

- All purrr functions are type-stable. They always return the advertised output type (e.g. `map()` returns lists; `map_dbl()` returns double vectors), or they throw an error.

- All `map()` functions either accept function, formulas (used for succinctly generating anonymous functions), a character vector (used to extract components by name), or a numeric vector (used to extract by position).

See the [iteration chapter](https://r4ds.had.co.nz/iteration.html){target="_blank"} of R for Data Science for further examples and details


# Efficient R programming {background-color="#552D62"}

## Growing Objects

Adding to an object in a loop, e.g. via `c()` or `cbind()`

```{r}
#| eval: false
res <- NULL
for (i in 1:10000) res <- c(res, 1)
```

forces a copy to be made at each iteration. **THIS IS BAD!**

It is far better to create an object of the necessary size first

```{r}
res <- numeric(10000)
for (i in seq_along(res)) res[i] <- 1
```

To initialise a list we can use
```{r}
res <- vector(mode = "list", length = 100)
```


## Benchmarking

There will usually be many ways to write code for a given task. To compare
alternatives, we can benchmark the expression

```{r}
#| cache: true
library(rbenchmark)
benchmark({res <- NULL;
           for (i in 1:10000) res <- c(res, 1)})$elapsed
benchmark({res <- numeric(10000)
           for (i in seq_along(res)) res[i] <- 1})$elapsed
```

Note the **BIG** difference between growing and initialising a vector (the latter around 150 times faster in this case).


## `for` loops revisited

Each loop has three components:

1. The output: allocate sufficient space before you start the loop

2. The sequence: this determines what you loop over

3. The body: the code that does the work

See <https://r4ds.had.co.nz/iteration.html#for-loops>

## Vectorization

Vectorization is operating on vectors (or vector-like objects) rather than individual elements.

Many operations in R are vectorized, e.g.

```{r}
x <- 1:3
y <- 3:1
x == y
log(x)
res <- list(a = 1:3, b = 1:6)
lengths(res)
```

We do not need to loop through each element!

## Recycling

Vectorized functions will recycle shorter vectors to create vectors of 
the same length
```{r}
1:4 + 0:1 # 1+0, 2+1, 3+0, 4+1
```
This is particularly useful for single values
```{r}
cbind(1, 3:4)
```
and for generating regular patterns
```{r}
paste0(rep(1:3, each = 2), c("a", "b"))
```

## Your turn!

1. Write a `for` loop to compute the mean of every column of in `mtcars`, saving each to a preallocated vector

2. Use `lapply()` with `rnorm` to generate a list of length 10 where the 1st item contains a vector of 1 sample from an $N(0,1)$ distribution, the 2nd item contains a vector of 2 samples from an $N(0,1)$ distribution up to the 10th item contains a vector of 10 samples from an $N(0,1)$ distibution.

3. Use `lapply()` with `rnorm` to generate a list of length 10, where the 1st item contains a vector of 5 samples from $N(1,1)$, the 2nd item contains a vector of 5 samples $N(2,1)$ and so on until you get 5 samples from $N(10,1)$

## Your turn! (sample solutions)

```{r}
#| eval: false

out <- numeric(ncol(mtcars))

for (i in seq_len(ncol(mtcars))) {
  out[i] <- mean(mtcars[[i]])
}
```

```{r}
#| eval: false
lapply(1:10, rnorm)
```

```{r}
#| eval: false
lapply(1:10, rnorm, n = 5)
```

## Vectorization and Matrices

Vectorizations applies to matices too, not only through matrix algebra
```{r}
M <- matrix(1:4, nrow = 2, ncol = 2)
M + M
```
but also vectorized functions
```{r}
M <- M + .3
round(M)
```

## Matrices and Recycling

Values are recycled down matrix, which is convenient for 
row-wise operations

```{r}
M <- matrix(1:6, nrow = 2, ncol = 3)
M
M - 1:2
```
To do the same for columns we would need to explicitly replicate, 
which is not so efficient.
```{r}
M - rep(1:3, each = 2)
```

## Vectorization vs For Loop

Operations that can be vectorized will be more efficient than a loop in R

```{r}
#| cache: true
M <- matrix(1:100000, nrow = 200, ncol = 500)
x <- 1:200
benchmark({for (i in 1:200){
             for (j in 1:500){
               M[i, j] <- M[i, j] - x[i]
             }
           }})$elapsed
benchmark({M - x})$elapsed
```

The latter is nearly 30 times faster!

## Row/Column-wise Operations

Several functions are available implementing efficient row/column-wise
operations, e.g. `colMeans()`, `rowMeans()`, `colSums()`, `rowSums()`, `sweep()`

```{r}
M <- matrix(1:4, nrow = 2, ncol = 2)
rowMeans(M)
```

These provide an alternative to iterating though rows and columns in R (the 
iteration happens in C, which is faster).

The **matrixStats** provides further "matricised" methods.

## Top tips for efficient programming

A **golden rule** in R programming is to access the underlying C/Fortran routines as quickly as possible; the fewer functions calls required to achieve this, the better. 

- Be careful never to grow vectors

- Vectorise code wherever possible

See [Efficient Programming](https://csgillespie.github.io/efficientR/programming.html){target="_blank"} for more details and examples.

# Writing functions {background-color="#552D62"}

## Components of a function

Functions are defined by three components:

 - the name of the function
 - the arguments of the function, inside `( )`
 - the body of the function that computes the result, inside `{ }`
 
They are created using `function()`

```{r}
t_statistic <- function(n) {
    x <- rnorm(n)
    y <- rnorm(n)
    t.test(x, y)$statistic
}
```

## Naming Functions

As with arguments, function names are important:

- use a name that describes what it returns (e.g. `t_statistic`) or what it does (e.g. `remove_na`)

- try to use one convention for combining words (e.g. snake case `t_statistic` or camel case `tStatistic`)

- avoid using the same name as other functions

## Specified Arguments

*specified* arguments are those named in the function definition, e.g. 
in `rnorm()`

```{r}
args(rnorm)
```
the arguments are `n`, `mean` and `sd`.

`mean` and `sd` have been given default values in the function definition, but `n` has not, so the function fails if the user does not pass a value to `n`

```{r, error = TRUE}
rnorm()
```

## Name and Order of Arguments

The user can pass objects to these arguments using their names or by supplying unnamed values in the right order

```{r}
rnorm(5, 1, 10)
rnorm(5, sd = 10)
```

So naming and order is important! Some guidelines

- put compulsory arguments first, e.g. data
- put rarely used arguments last, e.g. tolerance setting
- use short but meaningful argument names
- if relevant, use the same argument names as similar functions

## Using Arguments

Arguments are used as objects in the function code.

An new environment is created each time the function is called, separate 
from the global workspace.

```{r, error = TRUE}
x <- 1
y <- 3
f <- function(x, y){
    a <- 1
    x <- x + a
    x + y
}
f(x, y)
x
a
```

## Lexical Scoping

If an object is not defined within the function, or passed in as an argument, R looks for it in the *parent environment* where the function was defined

```{r, error = TRUE}
x <- 1
y <- 3
f <- function(x){
    x + y
}
f(x)
rm(y)
f(x)
```

It is safest (and best practice) to use arguments rather than depend on global variables! 

## Return Values (single)

By default, functions return the object created by the last line of code
```{r}
f <- function(x) {
    x <- x + 1
    log(x)
}
```

Alternatively `return()` can be used to terminate the function and return
a given object
```{r}
f <- function(x) {
    if (all(x > 0)) return(log(x))
    x[x <= 0] <- 0.1
    log(x)
}
```

## Return values (multiple)

Multiple objects can be returned in a list:

```{r}
mean_and_sd <- function(x) {
  res_mean <- mean(x, na.rm = TRUE)
  res_sd <- sd(x)
  
  list(mean = res_mean,
       sd = res_sd)
}

x <- 1:3
mean_and_sd(x)
```


## Your turn!

Write your own function, `variance`, to compute the variance of a numeric vector:

$$
Var(x) = \frac{1}{n-1}\sum_{i=1}^n(x_i - \bar{x})^2
$$

Make use of R's built in vectorisation.

Test it and compare your answer with the built-in `var()` function.

## Your turn! (a solution)

```{r}
#| eval: false

variance <- function(x) {
  1/(length(x) - 1) * sum((x - mean(x))^2)
}
```


# End matter {background-color="#552D62"}

## Resources

Material (very largely) inspired by and remixed from:

- Heather Turner and Erlap Dogu's [R Programming slides](https://www.heatherturner.net/teaching/turkey2018/r_programming/#1){target="_blank"} (with permission)

Additionally:

- [Efficient R, Chapter 3](https://csgillespie.github.io/efficientR/programming.html#the-apply-family) by Colin Gillespie and Robin Lovelace

- [R for Data Science](https://r4ds.had.co.nz/index.html){target="_blank"}, by Hadley Wickham and Garrett Grolemund, chapters on [iteration](https://r4ds.had.co.nz/iteration.html){target="_blank"} and [functions](https://r4ds.had.co.nz/functions.html){target="_blank"}

## License

Licensed under a Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License ([CC BY-NC-SA 4.0](https://creativecommons.org/licenses/by-nc-sa/4.0/){target="_blank"}).

# Extra stuff we probably won't get to {background-color="#552D62"}

(Use in Fuctions session of Advanced R course in Term 2 instead)

## Unspecified Arguments

`...` or the *ellipsis* allow unspecified arguments to be passed to the function. 

This device is used by functions that work with arbitrary numbers of objects, e.g. 
```{r}
args(sum)
sum(1, 4, 10, 2)
```

It can also be used to pass on arguments to another function, e.g.
```{r}
t_statistic <- function(x, g, ...) {
    t.test(x ~ g, ...)$stat
}
```

## Using `...`

Arguments passed to `...` can be collected into a
list for further analysis

```{r}
means <- function(...){
    dots <- list(...)
    vapply(dots, mean, numeric(1), na.rm = TRUE)
}
x <- 1
y <- 2:3
means(x, y)
```
Similarly the objects could be concatenated using `c()`

## Side Effects

A side-effect is a change outside the function that occurs when the
function is run, e.g.

- plot to the graphics window or other device
- printing output to the console
- write data to a file

A function *can* have many side-effects and a return value, but it is 
best practice to have a separate function for each task, e.g creating a 
plot or a table.

Writing to file is usually best done outside a function.
