---
title: Graphics
subtitle: R Foundations course
author: Ella Kaye | Department of Statistics | University of Warwick
title-slide-attributes:
  data-background-color: "#552D62"
date: 2022-11-21
date-format: long
format: 
  revealjs:
    theme: [default, ../../slides.scss]
    slide-number: true
    execute:
      echo: true
      code-overflow: wrap
---

## Overview

-   Plots in base R

-   ggplot2

-   Tables

# Plots in base R {background-color="#552D62"}

## Plots in RStudio

In RStudio, graphs are displayed in the Plots window. The plot is sized to fit the window and will be rescaled if the size of the window is changed.

Back and forward arrows allow you to navigate through graphs that have been plotted.

Graphs can be saved in various formats using the Export drop down menu, which also has an option to copy to the clipboard.

First we consider "no-frills" plots, for quick exploratory plots.

## palmerpenguins

```{r}
library(palmerpenguins)
```

## Boxplots

::: {.panel-tabset}
## Example 1

```{r}
boxplot(penguins$body_mass_g)
```


## Example 2
```{r}
with(penguins, boxplot(body_mass_g ~ species))
```
:::

## Histogram/Density

::: {.panel-tabset}
## Histogram

```{r}
hist(penguins$body_mass_g)
```

## Density
```{r}
plot(density(penguins$body_mass_g, na.rm = TRUE))
```
:::

## Scatterplots

::: {.panel-tabset}
## Vectors
```{r}
plot(1:10, 1:10)
```


## Data frame
```{r}
plot(bill_length_mm ~ bill_depth_mm, 
     data = subset(penguins, species == "Gentoo"))
```
:::

## Linear model diagnostic fits

```{r}
fit <- lm(bill_length_mm ~ bill_depth_mm, 
          data = penguins, subset = species == "Gentoo")

par(mfrow=c(2,2)) # see all four plots together
plot(fit)
```

# ggplot2 {background-color="#552D62"}

## Intro to ggplot2

```{r}
library(ggplot2)
```

```{r}
ggplot(data = NULL, aes(x = 1:10, y = 1:10)) +
  geom_point()
```

# Tables {background-color="#552D62"}

# End Matter {background-color="#552D62"}

## Resources

Material inspired by and remixed from:

- 

## License

Licensed under a Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License ([CC BY-NC-SA 4.0](https://creativecommons.org/licenses/by-nc-sa/4.0/){target="_blank"}).
